Map{R}educe in {MPI} for Large-scale Graph Algorithms

### 2010-05-04 paper/graph.tex
Minor
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@603 3fc257bb-8550-0410-81b3-a5c103da83f7
Modified pagerank description to make it more like other algorithms.
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@604 3fc257bb-8550-0410-81b3-a5c103da83f7

The PageRank algorithm assings a relative numeric rank to each
vertex in a graph.
It models the web as a directed graph $G(V,E)$, with each vertex $V_i
\in V$ representing a web page and each edge $(V_i, V_j) \in E$
representing a hyperlink from $V_i$ to $V_j$.  The probability of
moving from $V_i$ to another vertex $V_j$ is $\alpha/d_{out}(V_i) +
(1-\alpha)/|V|$, where $\alpha$ is a user-defined parameter (usually
0.8-0.9), $d_{out}(V_i)$ is the outdegree of vertex $V_i$, and $|V|$ is
the cardinality of $V$.  The first term represents the probability of
following a given link on page $V_i$; the second represents the
probability of moving to a random page.  For pages with no outlinks,
the first term is $\alpha/|V|$, indicating equal likelihood to move to
any other page.  Equivalently, the graph can be represented by a
matrix $A$~\cite{LangvilleMeyer05a}, with matrix entries $A_{ij} =
\alpha/d_{out}(V_i)$ if vertex $V_i$ links to $V_j$.  The PageRank
algorithm, then, is essentially a power-method iteration in which the
dominating computation is matrix-vector multiplication $A^T x=y$,
where $x$ is the PageRank vector from the previous iteration.
The algorithm in Figure \ref{fig:pr} performs these
iterations.

@@

%Input from R-MAT generator: \\
%\> MapReduce object $V$ containing all graph vertices:  Key=$V_i$, Value = NULL.  \\
%\> MapReduce object $E$ containing all graph edges: Key=$V_i$, Value = $V_j$. \\
%Initialize MapReduce object for matrix $A$  and \\
%MapReduce object $MT$ containing all vertices with no outlinks.\\
%\> Copy: $A = E$ \\
%\> Add: $V$ to $A$ \\
%\> Collate $A$: Vertex is Key \\
%\> Reduce $A$:  \\
%\> \> Input:  Key = $V_i$; MultiValue = NULL,$V_j$,$V_k$,... \\
%\> \> If (${nvalues} > 1$) \\
%\> \> \> Compute: $d_{out}(V_i) = {nvalues} - 1$. \\
%\> \> \> Output:  For each $V_j$ in MultiValue, Key = $V_i$; Value = ($V_j$, $\alpha/d_{out}(V_i)$). \\
%\> \> else \\
%\> \> \> Output to $MT$: Key = $V_i$; Value = 0. \\
%Initialize MapReduce object for pagerank vector $x$. \\
%\> Map:   \\
%\> \> Input:  Graph vertices $V$. \\
%\> \> Output:  For each vertex, Key = $V_i$; Value = $1/|V|$. \\
Input: \\
\> MapReduce object $A$ containing $|V| \times |V|$ matrix:  Key = $i$, Value = $[j, a_{ij}]$ \\
\> MapReduce object $MT$ containing indices of empty rows of $A$:  Key = $i$, Value = 0. \\
\> MapReduce object $x$ containing initial PageRank vector:  Key = $i$, Value = $1/|V|$. \\
While (${residual} > {tolerance}$) \\
\> Compute contribution for random jumps and zero-outdegree vertices. \\
\> \> Add: $x$ to $MT$. \\
\> \> Collate $MT$:  Vertex is Key. \\
\> \> Reduce $MT$:   \\
\> \> \> Input:  Key = $i$; MultiValue = $x_i, 0$ if $i \in {MT}$; MultiValue = $x_i$ if $i \notin {MT}$. \\
\> \> \> Compute:  Local contributions $cont$. \\
\> \> \> Output:  If ${nvalues} = 2$, Key = $i$; Value = 0. \\
\> \> MPI\_Allreduce with MPI\_SUM to compute global contribution $gcont$. \\
\> Compute $y = A^T x$.  \\
\> \> Copy $x$ to $y$. \\
\> \> Add $A$ to $y$. \\
\> \> Collate $y$:  Key is row index of $A$. \\
\> \> Reduce $y$: \\
\> \> \>  Input: Key = $i$; Multivalue = $i, [j, a_{ij}]$ for nonzeros $a_{ij}$ in $A$ \\
\> \> \>  Output:  Key = $j$; Value = $a_{ij} y_i$. \\
\> \> Collate $y$:  Key is column index of $A$. \\
\> \> Reduce $y$:  \\
\> \> \> Input:  Key = $j$; Multivalue = $a_{ij} y_i$ for all non-zero entries $a_{ij}$ of column $j$ \\
\> \> \> Output:  Key = $j$; Value = $\sum_i a_{ij} y_i$. \\
\> Add contribution $gcont$ to $y$; compute local max norm $norm$ of $y$. \\
\> \> Map $y$:  \\
\> \> \> Input:  Key = $i$, Value = $y_i$. \\
\> \> \> Compute: $y_i = y_i + gcont$; if $y_i > {norm}$, $norm = y_i$. \\
\> \> \> Output:  Key = $i$, Value = $y_i$. \\
\> \> MPI\_Allreduce $norm$ with MPI\_MAX to compute global max norm $gnorm$. \\
\> Scale $y$ by $gnorm$. \\
\> \> Map $y$: \\
\> \> \> Input:  Key = $i$, Value = $y_i$. \\
\> \> \> Output:  Key = $i$, Value = $y_i / {gnorm}$. \\
\> Compute residual. \\
\> \> Add $y$ to $x$. \\
\> \> Collate $x$:  Key is index $i$. \\
\> \> Reduce $x$:   \\
\> \> \> Input:  Key = $i$; Multivalue = $x_i, y_i$. \\
\> \> \> Compute:  if $|x_i - y_i| > {resid}$, ${resid} = |x_i - y_i|$ \\
\> \> MPI\_Allreduce with MPI\_MAX to compute $residual$. \\
\> $x = y$; \\
Output:  PageRank vector $x$.
  \end{tabbing}
 \end{minipage}}\end{center}

@@

%The MapReduce implementation performs two {\it map()} operations to
%initialize the graph matrix $A$ and PageRank vector $x$.  
The key kernel of PageRank is the matrix-vector multiplication $y = A^T x$.
In it, a {\it
collate()} operation gathers all row entries $a_{ij}$ with their
associated $x_i$ entry, and a {\it reduce()} computes $a_{ij} x_i$.  A
second {\it collate()} gathers, for each $j$, all contributions to the
column sum $\sum_i a_{ij} x_i$, which is computed by a second {\it
reduce()}.  MPI\_Allreduce calls are used to compute global norms and
residuals.

@@

Input: \\
$*$\> Pre-aggregated MapReduce object $A$ containing $|V| \times |V|$ matrix:  Key = $i$, Value = $[j, a_{ij}]$ \\
$*$\> Pre-aggregated MapReduce object $MT$ containing indices of empty rows of $A$:  Key = $i$, Value = 0. \\
$*$\> Pre-aggregated MapReduce object $x$ containing initial PageRank vector:  Key = $i$, Value = $1/|V|$. \\
While (${residual} > {tolerance}$) \\
\> Compute contribution for random jumps and zero-outdegree vertices. \\
\> \> Add: $x$ to $MT$. \\
$*$\> \> Convert $MT$:  Vertex is Key. \\
\> \> Reduce $MT$:   \\
\> \> \> Input:  Key = $i$; MultiValue = $x_i, 0$ if $i \in {MT}$; MultiValue = $x_i$ if $i \notin {MT}$. \\
\> \> \> Compute:  Local contributions $cont$. \\
\> \> \> Output:  If ${nvalues} = 2$, Key = $i$; Value = 0. \\
\> \> MPI\_Allreduce with MPI\_SUM to compute global contribution $gcont$. \\
\> Compute $y = A^T x$.  \\
\> \> Copy $x$ to $y$. \\
\> \> Add $A$ to $y$. \\
$*$\> \> Convert $y$:  Key is row index of $A$. \\
\> \> Reduce $y$: \\
\> \> \>  Input: Key = $i$; Multivalue = $i, [j, a_{ij}]$ for nonzeros $a_{ij}$ in $A$ \\
\> \> \>  Output:  Key = $j$; Value = $a_{ij} y_i$. \\
\> \> Collate $y$:  Key is column index of $A$. \\
\> \> Reduce $y$:  \\
\> \> \> Input:  Key = $j$; Multivalue = $a_{ij} y_i$ for all non-zero entries $a_{ij}$ of column $j$ \\
\> \> \> Output:  Key = $j$; Value = $\sum_i a_{ij} y_i$. \\
\> Add contribution $gcont$ to $y$; compute local max norm $norm$ of $y$. \\
\> \> Map $y$:  \\
\> \> \> Input:  Key = $i$, Value = $y_i$. \\
\> \> \> Compute: $y_i = y_i + gcont$; if $y_i > {norm}$, $norm = y_i$. \\
\> \> \> Output:  Key = $i$, Value = $y_i$. \\
\> \> MPI\_Allreduce $norm$ with MPI\_MAX to compute global max norm $gnorm$. \\
\> Scale $y$ by $gnorm$. \\
\> \> Map $y$: \\
\> \> \> Input:  Key = $i$, Value = $y_i$. \\
\> \> \> Output:  Key = $i$, Value = $y_i / {gnorm}$. \\
\> Compute residual. \\
\> \> Add $y$ to $x$. \\
$*$\> \> Convert $x$:  Key is index $i$. \\
\> \> Reduce $x$:   \\
\> \> \> Input:  Key = $i$; Multivalue = $x_i, y_i$. \\
\> \> \> Compute:  if $|x_i - y_i| > {resid}$, ${resid} = |x_i - y_i|$ \\
\> \> MPI\_Allreduce with MPI\_MAX to compute $residual$. \\
\> $x = y$; \\
Output:  PageRank vector $x$.
  \end{tabbing}
 \end{minipage}}\end{center}

@@

The initial step is to store a copy of the graph edges as key/value
(KV) pairs in a auxiliary MapReduce object $G_0$, for use later in the
algorithm.  The first {\it map()} operation converts edge keys to
vertex keys with edge values.  After the {\it collate()}, each vertex
has a list of vertices it is connected to; the first {\it reduce()}
can thus flag one vertex $V_i$ in each edge with a degree count $D_i$.
The second {\it collate()} and {\it reduce()} assign a degree count
$D_j$ to the other vertex in each edge.  In the third {\it map()},
only the lower-degree vertex in each edge emits its edges as key/value
(KV) pairs.  The task of the third {\it reduce()} is to emit
``angles'' for each of these low-degree vertices.  An ``angle'' is a
root vertex $V_i$, with two edges to vertices $V_1$ and $V_2$, i.e. a
triangle without the third edge $(V_1,V_2)$.  The {\it reduce()} emits
a list of all angles of vertex $V_i$, by a double loop over the edges
of $V_i$.  Note that the aggregate volume of KV pairs emitted at this
stage is minimized by having only the low-degree vertex in each edge
generate angles.

>>>>

The PageRank algorithm assings a relative numeric rank to each
vertex in a graph.
It models the web as a directed graph $G(V,E)$, with each vertex $V_i
\in V$ representing a web page and each edge $(V_i, V_j) \in E$
representing a hyperlink from $V_i$ to $V_j$.  The probability of
moving from $V_i$ to another vertex $V_j$ is $\alpha/d_{out}(V_i) +
(1-\alpha)/|V|$, where $\alpha$ is a user-defined parameter (usually
0.8-0.9), $d_{out}(V_i)$ is the outdegree of vertex $V_i$, and $|V|$ is
the cardinality of $V$.  The first term represents the probability of
following a given link on page $V_i$; the second represents the
probability of moving to a random page.  For pages with no outlinks,
the first term is $\alpha/|V|$, indicating equal likelihood to move to
any other page.  Equivalently, the graph can be represented by a
matrix $A$~\cite{LangvilleMeyer05a}, with matrix entries $A_{ij} =
\alpha/d_{out}(V_i)$ if vertex $V_i$ links to $V_j$.  To maintain the 
sparsity of $A$, terms for random jumps and zero out-degree vertices are
not explicitly stored in $A$.  Rather, they are computed separately
as adjustments to the PageRank vector.  The kernel of the PageRank
algorithm, then, is a power-method iteration consisting of 
matrix-vector multiplication $A^T x=y$,
where $x$ is the PageRank vector from the previous iteration, and
a few dot product and norm computations.

The algorithm in Figure \ref{fig:pr} performs these
iterations.  The nonzeros of matrix $A$ are initialized as described above;
the PageRank vector $x$ is initialized uniformly.  A MapReduce object $MT$
contains the indices of all-zero rows of $A$, corresponding to vertices
with no outlinks.  In step 1 of a PageRank interation, $MT$ is used 
in dot products to compute adjustments to the
PageRank vector to represent uniform probability of going from a leaf page
to any other page.  Adjustments for random jumps from any page are also
computed in this step.  Step 2, the matrix-vector product $A^T x$, is the most
expensive part of the computation, requiring two passes over the nonzero
structure of $A$.  In the first pass, {\it collate()} 
gathers all row entries $a_{ij}$ with their
associated $x_i$ entry, and {\it reduce()} computes $a_{ij} x_i$.  
In the second pass, 
{\it collate()} gathers, for each $j$, all contributions to the
column sum $\sum_i a_{ij} x_i$, which is computed by a second {\it reduce()}. 
Steps 3 and 4 adjust and scale the 
product vector; the work is proportional to $|V|$.  Step 6 computes the
residual to determine whether the PageRank vector has converged; work is
again proportional to $|V|$.  Throughout the algorithm, we exploit the
ability to call MPI\_Allreduce to compute global norms and residuals.

@@

%Input from R-MAT generator: \\
%\> MapReduce object $V$ containing all graph vertices:  Key=$V_i$, Value = NULL.  \\
%\> MapReduce object $E$ containing all graph edges: Key=$V_i$, Value = $V_j$. \\
%Initialize MapReduce object for matrix $A$  and \\
%MapReduce object $MT$ containing all vertices with no outlinks.\\
%\> Copy: $A = E$ \\
%\> Add: $V$ to $A$ \\
%\> Collate $A$: Vertex is Key \\
%\> Reduce $A$:  \\
%\> \> Input:  Key = $V_i$; MultiValue = NULL,$V_j$,$V_k$,... \\
%\> \> If (${nvalues} > 1$) \\
%\> \> \> Compute: $d_{out}(V_i) = {nvalues} - 1$. \\
%\> \> \> Output:  For each $V_j$ in MultiValue, Key = $V_i$; Value = ($V_j$, $\alpha/d_{out}(V_i)$). \\
%\> \> else \\
%\> \> \> Output to $MT$: Key = $V_i$; Value = 0. \\
%Initialize MapReduce object for pagerank vector $x$. \\
%\> Map:   \\
%\> \> Input:  Graph vertices $V$. \\
%\> \> Output:  For each vertex, Key = $V_i$; Value = $1/|V|$. \\
Input: \\
\> MapReduce object $A$ containing $|V| \times |V|$ matrix:  Key = $i$, Value = $[j, a_{ij}]$ \\
\> MapReduce object $MT$ containing indices of all-zero rows of $A$:  Key = $i$, Value = 0. \\
\> MapReduce object $x$ containing initial PageRank vector:  Key = $i$, Value = $1/|V|$. \\
While (${residual} > {tolerance}$) \\
\> 1. Compute contribution for random jumps and zero-outdegree vertices. \\
\> \> Add: $x$ to $MT$. \\
\> \> Collate $MT$:  Vertex is Key. \\
\> \> Reduce $MT$:   \\
\> \> \> Input:  Key = $i$; MultiValue = $x_i, 0$ if $i \in {MT}$; MultiValue = $x_i$ if $i \notin {MT}$. \\
\> \> \> Compute:  Local contributions $c_{local}$. \\
\> \> \> Output:  If ${nvalues} = 2$, Key = $i$; Value = 0. \\
\> \> MPI\_Allreduce $c_{local}$ with MPI\_SUM to compute global contribution $c_{global}$. \\
\> 2. Compute $y = A^T x$.  \\
\> \> Copy $x$ to $y$. \\
\> \> Add $A$ to $y$. \\
\> \> Collate $y$:  Key is row index of $A$. \\
\> \> Reduce $y$: \\
\> \> \>  Input: Key = $i$; Multivalue = $i, [j, a_{ij}]$ for nonzeros $a_{ij}$ in $A$ \\
\> \> \>  Output:  Key = $j$; Value = $a_{ij} y_i$. \\
\> \> Collate $y$:  Key is column index of $A$. \\
\> \> Reduce $y$:  \\
\> \> \> Input:  Key = $j$; Multivalue = $a_{ij} y_i$ for all non-zero entries $a_{ij}$ of column $j$ \\
\> \> \> Output:  Key = $j$; Value = $\sum_i a_{ij} y_i$. \\
\> 3. Add contribution $c_{global}$ to $y$; compute local max norm $n_{local}$ of $y$. \\
\> \> Map $y$:  \\
\> \> \> Input:  Key = $i$, Value = $y_i$. \\
\> \> \> Compute: $y_i = y_i + c_{global}$; if $y_i > n_{local}$, $n_{local} = y_i$. \\
\> \> \> Output:  Key = $i$, Value = $y_i$. \\
\> \> MPI\_Allreduce $n_{local}$ with MPI\_MAX to compute global max norm $n_{global}$. \\
\> 4. Scale $y$ by $n_{global}$. \\
\> \> Map $y$: \\
\> \> \> Input:  Key = $i$, Value = $y_i$. \\
\> \> \> Output:  Key = $i$, Value = $y_i / {n_{global}}$. \\
\> 5. Compute residual. \\
\> \> Add $y$ to $x$. \\
\> \> Collate $x$:  Key is index $i$. \\
\> \> Reduce $x$:   \\
\> \> \> Input:  Key = $i$; Multivalue = $x_i, y_i$. \\
\> \> \> Compute:  if $|x_i - y_i| > {resid}$, ${resid} = |x_i - y_i|$ \\
\> \> MPI\_Allreduce with MPI\_MAX to compute $residual$. \\
\> 6. $x = y$; \\
Output:  PageRank vector $x$.
  \end{tabbing}
 \end{minipage}}\end{center}

@@

@@

Input: \\
$*$\> Pre-aggregated MapReduce object $A$ containing $|V| \times |V|$ matrix:  Key = $i$, Value = $[j, a_{ij}]$ \\
$*$\> Pre-aggregated MapReduce object $MT$ containing indices of empty rows of $A$:  Key = $i$, Value = 0. \\
$*$\> Pre-aggregated MapReduce object $x$ containing initial PageRank vector:  Key = $i$, Value = $1/|V|$. \\
While (${residual} > {tolerance}$) \\
\> Compute contribution for random jumps and zero-outdegree vertices. \\
\> \> Add: $x$ to $MT$. \\
$*$\> \> Convert $MT$:  Vertex is Key. \\
\> \> Reduce $MT$:   \\
\> \> \> Input:  Key = $i$; MultiValue = $x_i, 0$ if $i \in {MT}$; MultiValue = $x_i$ if $i \notin {MT}$. \\
\> \> \> Compute:  Local contributions $c_{local}$. \\
\> \> \> Output:  If ${nvalues} = 2$, Key = $i$; Value = 0. \\
\> \> MPI\_Allreduce $c_{local}$ with MPI\_SUM to compute global contribution $c_{global}$. \\
\> Compute $y = A^T x$.  \\
\> \> Copy $x$ to $y$. \\
\> \> Add $A$ to $y$. \\
$*$\> \> Convert $y$:  Key is row index of $A$. \\
\> \> Reduce $y$: \\
\> \> \>  Input: Key = $i$; Multivalue = $i, [j, a_{ij}]$ for nonzeros $a_{ij}$ in $A$ \\
\> \> \>  Output:  Key = $j$; Value = $a_{ij} y_i$. \\
\> \> Collate $y$:  Key is column index of $A$. \\
\> \> Reduce $y$:  \\
\> \> \> Input:  Key = $j$; Multivalue = $a_{ij} y_i$ for all non-zero entries $a_{ij}$ of column $j$ \\
\> \> \> Output:  Key = $j$; Value = $\sum_i a_{ij} y_i$. \\
\> Add contribution $c_{global}$ to $y$; compute local max norm $n_{local}$ of $y$. \\
\> \> Map $y$:  \\
\> \> \> Input:  Key = $i$, Value = $y_i$. \\
\> \> \> Compute: $y_i = y_i + c_{global}$; if $y_i > {n_{local}}$, $n_{local} = y_i$. \\
\> \> \> Output:  Key = $i$, Value = $y_i$. \\
\> \> MPI\_Allreduce $n_{local}$ with MPI\_MAX to compute global max norm $n_{global}$. \\
\> Scale $y$ by $n_{global}$. \\
\> \> Map $y$: \\
\> \> \> Input:  Key = $i$, Value = $y_i$. \\
\> \> \> Output:  Key = $i$, Value = $y_i / {n_{global}}$. \\
\> Compute residual. \\
\> \> Add $y$ to $x$. \\
$*$\> \> Convert $x$:  Key is index $i$. \\
\> \> Reduce $x$:   \\
\> \> \> Input:  Key = $i$; Multivalue = $x_i, y_i$. \\
\> \> \> Compute:  if $|x_i - y_i| > {resid}$, ${resid} = |x_i - y_i|$ \\
\> \> MPI\_Allreduce with MPI\_MAX to compute $residual$. \\
\> $x = y$; \\
Output:  PageRank vector $x$.
  \end{tabbing}
 \end{minipage}}\end{center}

@@

The initial step is to store a copy of the graph edges as key/value
(KV) pairs in an auxiliary MapReduce object $G_0$, for use later in the
algorithm.  The first {\it map()} operation converts edge keys to
vertex keys with edge values.  After the {\it collate()}, each vertex
has a list of vertices it is connected to; the first {\it reduce()}
can thus flag one vertex $V_i$ in each edge with a degree count $D_i$.
The second {\it collate()} and {\it reduce()} assign a degree count
$D_j$ to the other vertex in each edge.  In the third {\it map()},
only the lower-degree vertex in each edge emits its edges as key/value
(KV) pairs.  The task of the third {\it reduce()} is to emit
``angles'' for each of these low-degree vertices.  An ``angle'' is a
root vertex $V_i$, with two edges to vertices $V_1$ and $V_2$, i.e. a
triangle without the third edge $(V_1,V_2)$.  The {\it reduce()} emits
a list of all angles of vertex $V_i$, by a double loop over the edges
of $V_i$.  Note that the aggregate volume of KV pairs emitted at this
stage is minimized by having only the low-degree vertex in each edge
generate angles.
